---
title: 'Project 2'
author: "Orlagh Keane,s2084384"
output:
  html_document:
    number_sections: yes
  pdf_document:
    number_sections: yes
header-includes:
  - \newcommand{\bm}[1]{\boldsymbol{#1}}
  - \newcommand{\mat}[1]{\begin{bmatrix}#1\end{bmatrix}}
---

```{r setup, include = FALSE}
# Modify this setup code chunk to set options
# or add extra packages etc if needed.
# See the project instructions for more details
# on what code to show, and where/how.

# Set default code chunk options
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE
)

suppressPackageStartupMessages(library(tidyverse))
theme_set(theme_bw())

# To give the same random number sequence every time the document is knit:ed,
# making it easier to discuss the specific numbers in the text:
set.seed(12345L)
```

```{r code=readLines("code.R"), eval=TRUE, echo=FALSE, results='hide'}
# Do not change this code chunk
# Load function definitions
source("code.R")
```

# Part 1: 3D printer
```{r code=readLines("code.R"), eval=TRUE, echo=FALSE, results='hide'}
# Load the filament1 dataset
load(file = "filament1.rda")
```

a) A function filament1_predict computes predictive distributions and 95% prediction intervals for a new dataset, the function returns a data.frame with variables mean, sd, lwr, and upr, summarizing the predictive distribution for each row of the new data. 

Using the function to compute probabilistic predictions of Actual_Weight using the two estimated models and the filament1 as new data. Using a level of significance of 5% for computing the prediction intervals, the predictions are represented visually together with the observed data.
```{r eval=TRUE, echo=TRUE}
load("filament1.rda")
# Predict using filament1_predict function
est_A <- filament1_estimate(data = filament1, model = "A")
est_B <- filament1_estimate(data = filament1, model = "B")

pred_filament1_A <- filament1_predict(est_A, newdata = filament1, model = "A")
pred_filament1_B <- filament1_predict(est_B, newdata = filament1, model = "B")

ggplot(rbind(cbind(pred_filament1_A, filament1, Model = "A"),cbind(pred_filament1_B, filament1, Model = "B")), mapping=aes(CAD_Weight))+
  geom_line(aes(y=pred_mean, col= Model)) +
  geom_ribbon(aes(ymin=lwr, ymax=upr, fill=Model), alpha=0.25) +
  geom_point(aes(y=Actual_Weight), data = filament1)

```




c) The function leave1out performs leave-one-out cross-validation for the selected model for each observation i = 1, . . . , N . In this function, it first estimates the model parameters using {(xj , yj ), j ̸= i}, computse the prediction information based on xi for prediction model Fi, and then computes the required scores S(Fi,yi). 

The output is a data.frame that extends the original data frame with four additional columns pred_mean, pred_sd, se and ds of leave-one-out prediction means, standard deviations, and prediction scores. leave1out obtains average leave-one-out scores S(Fi,yi) for each model and type of score for filament1 data. 

```{r eval=TRUE, echo=TRUE}
leave1out_A <- leave1out(filament1, "A")
leave1out_B <- leave1out(filament1, "B")

SE_data <- data.frame(CAD_Weight = leave1out_A$CAD_Weight, se_A = leave1out_A$se, se_B = leave1out_B$se)

ggplot() +
  geom_point(aes(CAD_Weight, se_A, colour = "A"), data = SE_data, alpha = 0.7) +
  geom_point(aes(CAD_Weight, se_B, colour = "B"), data = SE_data, alpha = 0.7) +
  labs(title = "Squared Error (SE) scores comparison", x = "CAD_Weight", y = "SE Score")

DS_data <- data.frame(CAD_Weight = leave1out_A$CAD_Weight, ds_A = leave1out_A$ds, ds_B = leave1out_B$ds)

ggplot() +
  geom_point(aes(CAD_Weight, ds_A, colour = "A"), data = DS_data, alpha = 0.7) +
  geom_point(aes(CAD_Weight, ds_B, colour = "B"), data = DS_data, alpha = 0.7) +
  labs(title = "Dawid-Sebastiani (DS) scores comparison", x = "CAD_Weight", y = "DS Score") 
```

The first graph, which represents the Squared Error scores, indicates that both models perform similarly for smaller CAD_Weights, with lower SE scores suggesting better accuracy. However, as the CAD_Weights increase, the SE scores for the two models diverge more significantly. This implies that the models' predictions are more accurate for lighter weights compared to heavier ones.

For the Dawid-Sebastiani scores, the plot reveals that the discrepancy between the two models is more pronounced across all CAD_Weights. Specifically, the DS score tends to increase as the CAD_Weights increase. The negative DS scores indicate that the observed data is not fitting the model's predictions well, possibly, the model is not a good fit for the data.


d) A Monte Carlo estimate of the p-value, testing the exchangeability between model predictions from A and B against the alternative hypothesis that B is better than A, and the Monte Carlo standard errors are presented below.    


```{r eval=TRUE, echo=TRUE}
score_diff <- data.frame(se = SE_data$se_A - SE_data$se_B, ds = DS_data$ds_A - DS_data$ds_B)
statistic0 <- score_diff %>% summarise(se = mean(se), ds = mean(ds))

J <- 10000
statistic <- data.frame(se = numeric(J), ds = numeric(J))

for (loop in seq_len(J)) {
  random_sign <- sample(c(-1, 1), size = nrow(score_diff), replace = TRUE)
  statistic[loop, ] <- score_diff %>% summarise(se = mean(random_sign * se), ds = mean(random_sign * ds))
}

p_values <- statistic %>%
  summarise(se = mean(se > statistic0$se), ds = mean(ds > statistic0$ds))
knitr::kable(p_values, caption = "Monte Carlo estimate of the p-value")

se_se <- sqrt((p_values$se * (1 - p_values$se)) / J)
se_ds <- sqrt((p_values$ds * (1 - p_values$ds)) / J)

knitr::kable(data.frame(se_se = se_se, se_ds = se_ds), caption = "Monte Carlo standard errors")

```
A high p-value for the SE score of 0.4989 suggests that the differences in predictions between models A and B are not statistically significant. This indicates that the two models are largely exchangeable with respect to the SE score.

A lower p-value for the DS score of 0.0436 suggests that the differences in predictions between models A and B are statistically significant. This implies that the models are not entirely exchangeable with respect to the DS score.

While the two scores present contrasting results, it might suggest that the observations are not best suited for a DS scoring method.

The standard errors for both SE and DS scores are relatively low, further supporting this statistical significance. Therefore, model B does appear to be better at predicting the outcomes compared to model A for the given data.



# Part 2: Archaeology in the Baltic sea

These are the approximate values for the Monte Carlo integration method.

```{r eval=TRUE, echo=TRUE}
result <- estimate(y = c(237, 256), xi = 1/1001, a = 0.5, b = 0.5, K = 10000)
knitr::kable(result)
```

The Monte Carlo integration estimated values that make some sense for the priors and the context. 
The estimate of 982 individuals is fairly close to the 1000 that the archaeologist first thought there would have been.

The estimated phi value of approximately 0.374 suggests that the probability of finding a femur is estimated to be around 37.4% which is much less than the archaeologist expected at 50%. 

The very low value of p_y of 0.0000079 indicates that the observed data is quite unlikely under the assumed prior distributions.

The evidence suggests that the actual values of N and phi might differ a lot from the initial assumptions, and therefore a different analysis might be better suited.



# Code appendix

```{r code=readLines("code.R"), eval=FALSE, echo=TRUE}
# Do not change this code chunk
```
